---
output: github_document
---

<!-- README.md is generated from README.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# dspline

<!-- badges: start -->
[![R-CMD-check](https://github.com/glmgen/dspline/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/glmgen/dspline/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/dspline)](https://CRAN.R-project.org/package=dspline)
<!-- badges: end -->

These are *not* B-splines:

```{r db-splines, echo=FALSE, dev="png", fig.width=7, fig.height=3, dpi=300}
library(dspline)
n = 50
k = 2

set.seed(11)
e = runif(n, -1/(2.5*(n+1)), 1/(2.5*(n+1)))
xd = 1:n/(n+1) + e
x = seq(0, 1, length = 5*n)
knot_idx = round(seq((k+1) + 5, (n-1) - 5, length = 4))
N1 = n_mat(k, xd, knot_idx = knot_idx)
N2 = n_eval(k, xd, x, knot_idx = knot_idx, N = N1)

par(mar = rep(0.01, 4))
matplot(x, N2, type = "l", lty = 1, col = 1:ncol(N2), 
        xlim = c(0, 1), ylim = c(0, 1), axes = FALSE)
matplot(xd, N1, type = "p", pch = 19, col = 1:ncol(N1), add = TRUE)
abline(v = xd[knot_idx], lty = 2, lwd = 0.5, col = 8)
```

These are called *discrete* B-splines. They span a function space called
**discrete splines**, which are analogous to splines, but defined in terms of 
a suitable discrete notion of smoothness. 

- Discrete splines have continuous *discrete* derivatives at their knots (rather
  than continuous derivatives, as splines do). 

- They have important computational properties, like the fact that interpolation
within the space of discrete splines can be done in *constant-time*. 

- They are intimately connected to trend filtering (they provide the basis
  representation that underlies the trend filtering estimator). 

For more background, see the monograph:
  ["Divided differences, falling factorials, and discrete splines:
  Another look at trend filtering and related
  problems"](https://www.stat.berkeley.edu/~ryantibs/papers/dspline.pdf).

The `dspline` package provides tools for computations with discrete splines. The
core routines are written in C++ for efficiency. See the
[reference index](https://glmgen.github.io/dspline/reference/index.html) for a
summary of the tools that are available.

## Installation

To install the released version from CRAN:

``` r
install.packages("dspline")
```

To install the development version from GitHub:

``` r
# install.packages("pak")
pak::pak("glmgen/dspline")
```
